---
title: "ANO habitat-species networks"
editor: visual
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

## Building habitat-species networks using ANO data

Ruben E. Roos, Joachim Töpper

## Running Code

### Interesting reading material

https://cran.r-project.org/web/packages/bipartite/vignettes/Intro2bipartite.pdf

### For fixing errors when reading Scandinavian letters

https://www.r-bloggers.com/2013/01/r-and-foreign-characters

https://community.rstudio.com/t/possible-bug-non-ascii-characters-not-displayed-properly-in-the-rstudio-release-2021-09-02/126312/2

```{r Adjust to deal with norwegian letters}
#|output: false
#|message: false
#|results: hide
Sys.setlocale(locale='no_NB.utf8')

#knitr::opts_chunk$set()
```

### Load required packages

```{r Load required packages}
#| echo: false
#| message: false
#| warning: false
#| results: hide

#load libraries
library(reshape)
library(vegan)
library(bipartite)
library(tidyverse)
library(randomcoloR)
library(visNetwork)

#Installs the github version (2.19) of bipartite, https://github.com/biometry/bipartite/issues/17
#library(devtools)
#install_github("biometry/bipartite/bipartite")

#Package bipartite needs to be manually installed and updated, can be downloaded here: https://github.com/biometry/bipartite/blob/master/bipartite_2.19.tar.gz, then use tools -> install packages, and tools -> check for package updates
```

```{r Import data}
#| echo: false
#| warning: false
#| output: false

#Read and review the data
#ANO.dat <- read.csv("~/Projects/ANO SATS/ANO_network_SATS/Data/ANO.dat.RR.csv", stringsAsFactors=TRUE)
#head(ANO.dat)
#str(ANO.dat)
#names(ANO.dat)
#unique(ANO.dat$ano_flate_id)

ANO.fjell <- read_csv("~/Projects/ANO SATS/ANO_network_SATS/Data/ANO.data.fjell.csv")#silence warning about missing colnames

```

Before we start our analyses, we can explore the data a bit. In addition, we can calculate some basic metrics.

```{r Data structure}
#| output: false

names(ANO.fjell)
str(ANO.fjell)
unique(ANO.fjell$ano_flate_id)#list the unique flate IDs

```

We can calculate some general metrics e.g., the average number of species per ANO-punkt, per NiN-type, etc.

```{r Basic metrics}

# Calculates the most frequently recorded nature type
ANO.fjell %>% 
  count(kartleggingsenhet_1m2) %>% 
  arrange(desc(n))#T3-C-2 (kalkfattig fjell-lynghei) is by far the most common nature type recorded

# Calculates number of species per ANO point
sr <- ANO.fjell %>% 
  group_by(ano_punkt_id) %>% 
  summarise(species_richness = n(), nature_type = first(kartleggingsenhet_1m2))
sr %>% head()

# Calculate average number of species per NiN nature type
sr %>% 
  group_by(nature_type) %>% 
  summarise(mean_sr = mean(species_richness)) %>% 
  arrange(desc(mean_sr))#orders to view the most species rich nature type first

# Calculate hwo frequently the redlist categories occur per nature type
ANO.fjell %>% 
  group_by(kartleggingsenhet_1m2, Kategori.2021) %>% 
  tally()
  spread(value, n, fill = 0)

#select unique species per nature type
ANO.fjell %>% 
  select(kartleggingsenhet_1m2, Species, Kategori.2021) %>% 
  group_by(kartleggingsenhet_1m2) %>% 
  distinct(Species, .keep_all = TRUE) %>% #creates a species list per NiN nature type and their redlist status 
  summarise("Number of species" = n(),
    across(Kategori.2021, list(LC = ~ sum(. == "LC"),
                                DD = ~ sum(. == "DD"),
                                NT  = ~ sum(. == "NT"), 
                               VU = ~ sum(. == "VU"),
                               EN = ~ sum(. == "EN"),
                               CR = ~ sum(. == "CR"),
                               RE = ~ sum(. == "RE")))) %>% 
  rowwise() %>% 
  mutate(RL = sum(Kategori.2021_DD, Kategori.2021_NT, Kategori.2021_VU, 
                    Kategori.2021_EN, Kategori.2021_CR, Kategori.2021_RE), 
         TS = sum(Kategori.2021_VU, Kategori.2021_EN, Kategori.2021_CR)) %>% #add number of readlisted species (DD-RE), and threatened species (VU, EN, CR)
  arrange(desc(RL))#order nature types by most redlisted species: T27-C-6 (kalkfattig og intermediaer rabbepreget blokkmark) has most RL species


```

Right now, the dataset is very large, and for the proof of concept, we can subset five ANO-flater (sites) that are in mountains. Here we just select the five first in the dataset: ANO0420, ANO0184, ANO0101, ANO0879, and ANO1455

```{r Subset data if needed}

#list <- c("ANO0420", "ANO0184", "ANO0289", "ANO0879", "ANO1455")

#df <- ANO.dat %>% 
#  filter(ano_flate_id %in% list)

```

We now want to convert this dataset to a long format.

```{r Convert to long format}

ANO.fjell.long <- pivot_longer(cols = `Festuca.ovina`:`Erigeron.acris`, 
    names_to = "Species",
    values_to = "species_cover")

```

More than one rute (plot) per flate (site) can have the same nature/habitat type. It is probably good to save this "resolution". Optionally, we can take the mean for each nature/habitat type within each flate (site). We then select one flate (site) to visualize a network.

```{r}
#| echo: false

df <- df %>%
  pivot_longer(cols = `Festuca.ovina`:`Erigeron.acris`, 
    names_to = "Species",
    values_to = "species_cover") %>%  #convert to long format
  dplyr::rename(habitat_type = kartleggingsenhet_1m2, 
                site_id = ano_flate_id, 
                patch_id = ano_punkt_id) %>% 
  filter(!is.na(species_cover)) %>% 
  filter(!(species_cover == 0))  #remove species without cover (one species was recorded without cover is the plot did not have any vascular vegetation)
#These lines can be used if we want to aggregate species cover across all similar nature types within a site (ANO-flate)
#  group_by(site_id, habitat_type, Species) %>% 
#  summarize(mean_cover = mean(species_cover, na.rm = FALSE))


#subset one site and the columns we need
site1 <- subset(df, site_id == "ANO0184")
site1

#create the network in matrix format
ntwANO <- cast(site1, patch_id ~ Species, fun.aggregate = "mean", value = "species_cover")
#we can remove the first column with rownames
ntwANO_ <- ntwANO[,-1]
ntwANO <- as.matrix(ntwANO_)
colnames(ntwANO) <- colnames(ntwANO_)
#colnames(ntwANO) <- gsub(pattern = "Bombus_", replacement = "B. ", 
#                  x = colnames(ntwANO_), ignore.case = T)
#create a patch dataframe to store its properties.
#RR: original code for color doesn't work for all sites, as the number of habitats differs between them - needs a tidier solution, see below
patch <- unique(site1[, c("patch_id", "habitat_type")])
patch <- patch %>% 
  dplyr::group_by(habitat_type) %>% 
  dplyr::mutate(Color = randomcoloR::randomColor()) %>%
  dplyr::ungroup()

#patch$color <- c("gold", "green", "darkgreen", "orange", "lightgreen")
rownames(ntwANO) <- patch$patch_id
#create a vector of plants
plants <- cast(site1, Species ~ ., fun.aggregate = sum, value = "species_cover")
colnames(plants)[2] <- "cover"
plants$labs <- colnames(ntwANO)

```

### Network visualization

```{r}

#We can use package _bipartite_ to plot it in two different ways:

#prepare legend
legend <- unique(patch[,c("habitat_type", "Color")])
par(xpd = T) #allow plotting outside the plot
plotweb(ntwANO, col.low = as.character(patch$Color))
legend(x=0, y=0.25, as.character(legend$habitat_type), pch=21,
       col="#777777", pt.bg=as.character(legend$Color), 
       pt.cex=1, cex=.6, bty="n", ncol=2)
visweb(ntwANO, prednames = T, preynames = T, labsize = 0.6)

#See also: https://briatte.github.io/ggnet/
```

```{r}
#or with pretier tools:
library(igraph)
#prepare the data for igraph
links <- site1[,c("Species", "patch_id", "species_cover")]
colnames(links)[3] <- "weight"
node1 <-  unique(site1[,c("patch_id", "habitat_type")])
colnames(node1) <- c("node", "attribute")
node1$type <- "habitat"
site1 <- as.data.frame(site1)
node2 <-  data.frame(node = unique(site1[,c("Species")]), 
                     attribute = NA,
                     type = "species")
nodes <- rbind(node1, node2)
#create igraph object
net <- graph_from_data_frame(d=links,
                             vertices=nodes, directed=F) 
# Generate colors based habitat: 
clrs <- data.frame(nod = V(net)$attribute,
                   cols = c(patch$Color, rep("blue",44 )))
V(net)$color <- as.character(clrs$cols)
# Compute node degrees (#links) and use that to set node size:
deg <- degree(net, mode="all")
V(net)$size <- deg*3
# Setting them to NA will render no labels:
#V(net)$label <- as.character(nodes$id) #RR: this throws an error
# Set edge width based on weight:
E(net)$width <- E(net)$weight/3
#change arrow size and edge color:
E(net)$arrow.size <- .2 #but note no arrows in Unidirected graphs like this
E(net)$edge.color <- "gray80"
#prepare colors
cl <- unique(clrs)
cl$nod <- as.character(cl$nod)
cl$nod[which(is.na(cl$nod))] <- "Species"
plot(net, vertex.label = NA) #force vertex label NA to make visualization clearer.
legend(x=-1.5, y=-1.1, cl$nod, pch=21,
       col="#777777", pt.bg=as.character(cl$cols), 
       pt.cex=2, cex=.8, bty="n", ncol=2) #RR: doesn't work
```

There is one more package *visNetwork* that makes impressive interactive plots that may be better to visualize and explore these kind of graphs.

```{r}
#| eval: false


colnames(nodes)[1] <- "id"
nodes$shape <- "dot"  
nodes$shadow <- TRUE # Nodes will drop shadow
nodes$attribute <- as.character(nodes$attribute) 
nodes$attribute[10:23] <-  as.character(nodes$id[10:23])#this is horrible coding
nodes$title <- nodes$attribute # Text on click
nodes$label <- nodes$type # Node label
nodes$size <- deg*3 # Node size
nodes$borderWidth <- 2 # Node border width

nodes$color.background <- clrs$cols
nodes$color.border <- "black"
nodes$color.highlight.background <- "orange"
nodes$color.highlight.border <- "darkred"
links$width <- links$weight # line width
links$color <- "gray"    # line color  
#links$arrows <- "middle" # arrows: 'from', 'to', or 'middle'
links$smooth <- TRUE    # should the edges be curved?
links$shadow <- FALSE    # edge shadow
colnames(links)[1:2] <- c("from", "to")
visNetwork(nodes, links) 
```

### Examine network structure

How is this network structured? To calculate it's *nestedness*, we use weighted NODF, one of the more popular and robust nestedness metrics (Almeida-Neto et al. 2008, Almeida-Neto et al. 2010), but other options are also available.

```{r}
(obs <- networklevel(web = ntwANO, index = "weighted NODF"))
```

To know if `r obs` is more nested than expected by chance for this network, we need to compare it with a null model:

```{r null, message=FALSE, warning=FALSE}
nm <- nullmodel(web = ntwANO, N=100, method="vaznull")
null <- unlist(sapply(nm, networklevel, index="weighted NODF")) 
plot(density(null), xlim=c(min(obs, min(null)), max(obs, max(null))), 
		main="comparison of observed with null model Patefield")
abline(v=obs, col="red", lwd=2)    
praw <- sum(null>obs) / length(null)

```

Here, we can see that this network is less nested than expected by chance (p value `r ifelse(praw > 0.5, round(1-praw,3), round(praw, 3))`). This suggests that these species do not use habitats in a nested way (i.e. species-rich habitats do not host species found in species poor habitats) and we should check for other network structural features.

To gain further insights on the structure of species habitats networks, we can use another important structural metric: modularity. In this example, we will calculate a quantitative version of the metric (Dormann et al., 2014).

For info on this: Detecting modules in quantitative bipartite networks: the QuaBiMo algorithm -- arXiv Vanity (arxiv-vanity.com)

```{r}
#| fig.height: "160%"

res <- metaComputeModules(ntwANO, N = 5)
plotModuleWeb(res, displayAlabels = TRUE, displayBlabels = TRUE)

```

```{r}
#listModuleInformation(res)
#printoutModuleInformation(res)
modules.nulls <- sapply(nm, computeModules)
like.nulls <- sapply(modules.nulls, function(x) x@likelihood)
praw <- sum(like.nulls > res@likelihood) / length(like.nulls)
```

The network is more modular than expected by chance (p-value \< `r ifelse(praw > 0.5, round(1-praw,3), round(praw, 3))`). This confirms that each habitat tends to harbour a quite unique assemblage of species, with limited exchange between habitat patches. For the purpose of this worked example, we now focus on the description of these modules. We can see that patch 44, 51, 46, 24, 53, 62, 22, 66, etc...

```{r}
#we can calculate 2 values for each node
cz <- czvalues(res, weighted = TRUE, level = "lower")
#c : among-module connectivity
#z : within-module connectivity
#Olesen et al. (2006) give critical c and z values of 0.62 and 2.6, respectively. Species exceeding these values are deemed connectors or hubs of a network. The justification of these thresholds remains unclear to me. They may also not apply for the quantitative version.
plot(cz[[1]], cz[[2]], pch=16, xlab="c", ylab="z", 
     cex=0.8, las=1, col = patch$Color)
text(cz[[1]], cz[[2]], names(cz[[1]]), pos=4, cex=0.7)


#What does this mean?
```

Next, we can ask which habitat patches would we prioritize for its conservation value. For this we can calculate its strength (Bascopmte et al 2006).

```{r}
#we transpose the matrix to calculate it for the lower level.
patch$strength <- bipartite::strength(t(ntwANO), type="Bascompte")
patch

#produces NaN for strength??
```

### Looping through multiple sites at once

Now we want to loop through several sites at once. For start, we can use the five sites we used earlier.

```{r}
#| eval: false

unique(df$site_id)

source("toolbox.R") #load code developed by Song et al. and available in his paper.
#let's loop through all sites
#first we create empty objects to store the data
sites <- unique(df$site_id)
ntwks <- list()
nested <- c()
NODF <- c()
st_NODF <- c()
rob_rand <- c()
rob_real <- c()
rich <- c()
seminat <- c()
for(i in 1:length(sites)){
  sitex <- subset(df, site_id == sites[i])
  #create the network in matrix format
  ntwx <- cast(sitex, patch_id ~ Species, fun.aggregate = "sum", 
               value = "Abundance")
  #we can remove first column with rownames
  ntwx <- ntwx[,-1]
  #let's visualize it with bipartite
  #plotweb(ntwx)
  ntwks[[i]] <- ntwx
  #calculate nestedness
  nested[i] <- networklevel(web = ntwx, index = "weighted NODF")
  rob_rand[i] <- robustness(second.extinct(web = ntwx, nrep = 50, participant = "lower", method = "random"))
  NODF[i] <- nestedness_NODF(as.matrix(ntwx))
  st_NODF[i] <- comb_nest(web = ntwx, NODF = NODF[i], max_NODF = max_nest(ntwx))
  #reate a realistic extinction sequence
  
  ext_seq <- unique(sitex[,c("Patch", "Habitat")])
  #quick and dirty way to order habitats
  levels(ext_seq$Habitat) <- c("gCorridor", "bForest",
                               "cForest_grassland_boundary",
                               "fMaintained_drain", 
                               "eMaintained_roadside",
                               "dNon_flowering_crop_edge",
                               "aSemi_natural_grasslands")
  ext_seq$Patch <- order(as.character(ext_seq$Habitat))
  rob_real[i] <- robustness(second.extinct(web = ntwx, participant = "lower", method = "external", ext.row = ext_seq$Patch )) #garsslands first, forest, etc...
  rich[i] <- ncol(ntwx)
  seminat[i] <- length(subset(ext_seq, Habitat %in%
                                c("aSemi_natural_grasslands",
                                  "bForest"))$Patch)
}
sites_measures <- data.frame(sites, nested, NODF, st_NODF, rob_rand, rob_real, rich, seminat)





```

The `echo: false` option disables the printing of code (only output is displayed).

Calculate lots of things on the whole network

```{r}
networklevel(ntwANO)
```

Test Robustness (area of the area underneath the extinction curve) using random method (i.e. removing habitats randomly:

```{r}
ntw.ext.rand <- second.extinct(web = ntwANO, nrep = 100, participant = "lower", method = "random", details = FALSE)

ntw.ext.abun <- second.extinct(web = ntwANO, nrep = 50, participant = "lower", method = "abundance", details = FALSE)

ntw.ext.higher <- second.extinct(web = ntwANO, nrep = 100, participant = "higher", method = "abundance", details = FALSE)

#Calculate robustness (surface area underneath the extinction curve)
robustness(second.extinct(web = ntwANO, nrep = 50, participant = "lower", method = "abundance", details = TRUE))
```

Plot the extinction curve

```{r}
slope.bipartite(ntw.ext.rand, plot.it = TRUE)
slope.bipartite(ntw.ext.abun, plot.it = TRUE)
slope.bipartite(ntw.ext.higher, plot.it = TRUE)
```
